\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2009-10-29 honors_thesis/thesis/paper.tex
Revisions to my honors thesis, a bunch of other random stuff

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait, guess which way the branch will go and start
executing.  If the processor guessed right, execution is much farther ahead
than if it had just idly waited on the branch.  If it guessed wrong, execution
is right where it would have been anyway.  This is called speculative
execution or branch prediction.

@@

Because the motion of particles in PSO depends only on the \emph{positions} of
the particle and its neighbors, not the \emph{value} of the objective function
at those positions, speculative evaluation is possible.  The relative values of
the positions of the particle's neighbors do matter, in that the best one is
used as the neighborhood best position, but it's which position has the highest
value, not what that value is, that matters.  We then propose a new method of
parallelizing PSO in which additional processors are used to perform more
iterations, not add particles to the swarm.

Because we do not have to wait for a particle's position to be evaluated to know
where it might move to next, we can compute the set of all possible next
positions and send them to be evaluated along with the particle's current
position.  Once the current position of each particle has been evaluated, we
can determine which of the speculatively evaluated positions was correct.  We
have thus evaluated two iterations at once.  This can easily get unwieldy if
the set of next possible positions for each particle is large, but a wise
choice of topology limits that set to a reasonable size.

@@

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once.  In \sec{extra}, we discuss a variant of the
speculative algorithm which uses all of the speculative evaluations performed
instead of throwing away all but one for each particle.  We then show in
\sec{results} a comparison between the original PSO algorithm, our speculative
algorithm, and its variant.

@@

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  We use the Ring topology, where each particle
is neighbors only with two other particles, and a parallel adaptation of the
Complete topology~\cite{mcnabb-cec09}, where each particle is neighbors with
all other particles.  These are the two most commonly used topologies in the
literature and have been shown to work well in practice~\cite{bratton-sis07}.

@@

This second method only requires one round of passing information because
information about iteration $t$ and $t+1$ are passed at the same time.  Each
processor reconstructs the from the messages it receives all of the information
that it needs about its neighbors.  Messages are passed directly after
evaluating each particle and its children, so all messages are of the form of
$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
treated specially, so each particle can produce its initial set of speculative
children--neighbors need only pass their initial position.  This kind of
message passing necessitates the careful use of random seeds, so that when each
processor computes the motion equations for its neighbors it gets the same
results as its neighboring processors.

@@

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, except for the neighbor particles instead of the particle itself.

@@

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  However, if speculative
evaluation were not being performed, it is likely that the Complete topology
would be used.  To be fair in our comparisons, for the standard PSO algorithm
we compare to both the Random topology with two neighbors and to the Complete
topology.  All functions have 20 dimensions, and all results shown are averages
over 20 runs.  For ease of labeling, we call our speculative algorithm SEPSO,
the variant SEPSO Pick Best, and standard PSO just PSO, with the topology
following if necessary.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for a swarm size of 50
particles with SEPSO, which has been recommended as a good standard swarm
size~\cite{bratton-sis07}.  One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed.

@@

Lastly we look at functions that don't show much or any improvement with this
method.  The Ackley function is defined as $f(\Vec{x}) = -20\exp\left\{
-0.2\sqrt{\frac{1}{D}\sum_{i=1}^Dx_i^2}\right\} - \exp\left\{\frac{1}{D}
\sum_{i=1}^D\cos(2\pi x_i)\right\}+20+e$. \fig{ackley} shows a comparison of
our speculative algorithms versus the original PSO with swarm sizes of 200 and
800, respectively.  There is no significant difference between the performance
of SEPSO and PSO Complete, though the performance of SEPSO Pick Best is
interesting to note.  It quickly converged on a local optimum and compared to
the other algorithms had dismal performance on this function.

@@

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration, and it has been shown that for a 20-dimensional Griewank function,
400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  In fact, the
class of functions mentioned above could be considered a special case of these
functions, where ``enough'' exploration can be achieved with only a very small
number of particles proportional to the dimensionality of the problem.  For
these functions, one should use processors to increase the swarm size until
enough exploration has been reached, then use additional processors to increase
the number of iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  There are an infinite number of
local minima in the Rastrigin function, and it has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been done.  For
those functions the correct use of processors depends cannot be easily
prescribed and depends largely on the desired outcome.  It is likely that the
best thing to do with those functions is to use all processors to add particles
to the swarm and do no speculative evaluation.

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.  For
example, our technique required speculatively evaluating every possible next
position of each particle.  The number of speculative evaluations could
potentially be greatly reduced by using statistics for which branch of motion
particles most often follow.  Only the subset of the most probable next
locations would then be evaluated.  Using an asynchronous PSO variant would
also be helpful in this case, allowing particles to be at different iterations
in case the correct branch was not guessed, and needs to be evaluated
separately.

@@

Large parallel clusters are often required to successfully optimize practical
modern problems.  To properly use PSO with such clusters, a balance needs to be
made between using processors to increase the swarm size and using them to
increase the speed of the algorithm, or perhaps as yet unconsidered techniques.
This work is a first step in that direction.

>>>>

To solve this problem, we take inspiration from speculative execution
techniques commonly used in processors.  Modern processors, when faced with a
branch on which they must wait, guess which way the branch will go and start
executing.  If the processor guessed right, execution is much farther ahead
than if it had just idly waited on the branch.  If it guessed wrong, execution
restarts right where it would have been anyway.  This is called speculative
execution or branch prediction.

@@

Because the motion of particles in PSO depends only on the \emph{positions} of
the particle and its neighbors, not the \emph{value} of the objective function
at those positions, speculative evaluation is possible.  We do not have to wait
for a particle's position to be evaluated to know where it might move to next,
so we can compute the set of all possible next positions and send them to be
evaluated along with the particle's current position.  Once the current
position of each particle has been evaluated, we can determine which of the
speculatively evaluated positions matches the branch taken by the particle.  We
have thus evaluated two iterations at once.  This can easily get unwieldy if
the set of next possible positions for each particle is large, but a wise
choice of topology limits that set to a reasonable size.

@@

\sec{pso} describes the particle swarm optimization algorithm.  \sec{sepso}
then describes how speculative evaluation can be done in parallel PSO to
perform two iterations at once.  In \sec{extra}, we discuss a variant of the
speculative algorithm which uses all of the speculative evaluations performed
instead of throwing away all but one for each particle.  We then show in
\sec{results} a comparison between the original PSO algorithm, our speculative
algorithm, and its variant.  In \sec{conclusion} and \sec{future} we conclude
and discuss future work.

@@

There are many ways of defining the neighbors of a particle.  Different
neighborhood topologies have varying performance on different kinds of
functions, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.  An in-depth discussion of topologies in PSO is
beyond the scope of this paper.  The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In this paper we use those topologies, as well
as a parallel adaptation of the Complete topology called Random that has been
shown to simulate the behavior of Complete with far less
communication~\cite{mcnabb-cec09}.  Both the Ring and the Random topologies can
be defined as having $k$ neighbors; we use 2 neighbors for all of our
experiments.

@@

This second method only requires one round of passing information because
information about iteration $t$ and $t+1$ are passed at the same time.  Each
processor reconstructs from the messages it receives all of the information
that it needs about its neighbors.  Messages are passed directly after
evaluating each particle and its children, so all messages are of the form of
$\nonbest{\p}_t$ or $\nonbest{\s}_{t+1}$.  The first iteration needs to be
treated specially, so each particle can produce its initial set of speculative
children--neighbors need only pass their initial position.  This kind of
message passing necessitates the careful use of random seeds, so that when each
processor computes the motion equations for its neighbors it gets the same
results as its neighboring processors.

@@

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, except for the neighbor particles instead of the particle itself.  
\fig{specparticles} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

\begin{figure}
  \centering
  \includegraphics[width=.6\columnwidth]{messages2.eps}
  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
  and $\nset_{t+1}$ are both produced as in \fig{origparticle}.}
  \label{fig:specparticles}
\end{figure}

@@

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  However, if speculative
evaluation were not being performed, it is likely that the Complete topology
would be used.  To be fair in our comparisons, for the standard PSO algorithm
we compare to both the Random topology with two neighbors and to the Complete
topology.  All functions have 20 dimensions, and all results shown are averages
over 20 runs.  Except where noted, we use swarms of close to 50 particles, as
that has been recommended as a general-purpose swarm size~\cite{bratton-sis07}.
When the speculative swarms have 50 particles, regular PSO swarms have 200.
For ease of labeling, we call our speculative algorithm SEPSO, the variant
SEPSO Pick Best, and standard PSO just PSO, with the topology following if
necessary.

@@

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for a swarm size of 50
particles with SEPSO.  One can see in the figure that both SEPSO algorithms
initially outperformed PSO, but got stuck in a local optima and were eventually
outperformed.

@@

Lastly we look at functions that do not show much or any improvement with this
method.  The Ackley function is defined as $f(\Vec{x}) = -20\exp\left\{
-0.2\sqrt{\frac{1}{D}\sum_{i=1}^Dx_i^2}\right\} - \exp\left\{\frac{1}{D}
\sum_{i=1}^D\cos(2\pi x_i)\right\}+20+e$. \fig{ackley} shows a comparison of
our speculative algorithms versus the original PSO with swarm sizes of 50 and
200, respectively.  There is no significant difference between the performance
of SEPSO and PSO Complete, though the performance of SEPSO Pick Best is
interesting to note.  It quickly converged on a local optimum and compared to
the other algorithms had dismal performance on this function.

@@

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration, and it has been shown that for a 20-dimensional Griewank function,
400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  For these
functions, one should use processors to increase the swarm size until enough
exploration has been reached, then use additional processors to increase the
number of iterations performed.

There are still other functions for which it seems there is never enough
exploration, such as the Rastrigin function.  It has been shown that up to 4000
particles there is no point at which ``enough'' exploration has been
done~\cite{mcnabb-cec09}.  For those functions the correct use of processors
cannot be easily prescribed and depends largely on the desired outcome.  It is
likely that the best thing to do with those functions is to use all processors
to add particles to the swarm and do no speculative evaluation.

@@

This work represents the first attempt of which we are aware at using
additional processors in parallel with PSO to do something other than either
add particles to the swarm or parallelize the evaluation of the objective
function.  We anticipate that many future advancements can be made.  For
example, our technique required speculatively evaluating every possible next
position of each particle.  The number of speculative evaluations could
potentially be greatly reduced by using statistics for which branch of motion
particles most often follow.  Only the subset of the most probable next
locations would then be evaluated.  Using an asynchronous PSO variant would
also be helpful in this case, allowing particles to be at different iterations
in case the correct branch was not guessed and needs to be evaluated
separately.

@@

Large parallel clusters are often required to successfully optimize practical
modern problems.  To properly use PSO with such clusters, a balance needs to be
made between using processors to increase the swarm size and using them to
increase the speed of the algorithm, or for perhaps as yet unconsidered
techniques.  This work is a first step in that direction.
